Experiences with planning for natural language generation

### 2008-06-09 trunk/Papers/experiences/domains.tex
added some text for the crisp domain

>>>>

Sentence generation is the problem of computing, from a grammar and a
semantic representation, a single sentence that expresses this piece
of meaning.  This problem is traditionally split into several steps
\cite{reiter00building}.  In a first step, called \emph{sentence
  planning}, the semantic representation is first enriched with more
information; for instance, \emph{referring expressions}, which refer
to individuals that we want to talk about, are determined at this
point.  In a second step, called \emph{surface realization}, this
enriched representation is then translated into a natural-language
sentence, using the grammar.

However, determining referring expressions (REs) and realization
interact, so it would be beneficial to perform both steps together.
This was the goal of the SPUD system \cite{Stone2003a}, which
performed both steps together using a top-down generation algorithm
based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
entries were equipped with semantic and pragmatic information.
Unfortunately, SPUD suffered from having to explore a huge search
space, and had to resort to a non-optimal greedy search strategy to
retain reasonable efficiency.

\cite{KolSto07} attempt to improve the efficiency by translating the
SPUD sentence generation problem into a planning problem and using a
planning algorithm for generation.  We will now illustrate how they do
this by a (simplified) example.

\begin{figure}
  \centering
  xxxx
  \caption{The example grammar.}
  \label{fig:white-rabbit-sleeps-grammar}
\end{figure}

\begin{figure}
  \centering
  xxxx
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

Consider a knowledge base containing the individuals $r_1$ and $r_2$;
$r_1$ and $r_2$ are rabbits, $r_1$ is white and $r_2$ is brown, and
$r_1$ sleeps.  Let's say we want to express the information
$\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar shown in
Fig.~\ref{fig:white-rabbit-sleeps-grammar}.  This grammar consists of
\emph{elementary trees}, each of which contributes certain
\emph{semantic content}.  We can instantiate these trees by
substituting individuals for \emph{semantic roles}, such as
$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree
instances as shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to
obtain the sentence ``The white rabbit sleeps''.  The SPUD algorithm
computes this grammatical derivation by starting with the elementary
tree for ``sleeps''; this satisfies the need to convey the semantic
information, but introduces a need to generate a noun phrase (NP) for
the subject, which also refers uniquely to the target referent $r_1$.
In a second step, it substitutes the tree for ``the rabbit'' into the
open NP leaf, which makes the derivation grammatically complete; but
as there are two different individuals that could be described as
``the rabbit'' -- technically, $r_2$ is still a \emph{distractor} for
the referring expression --, we are still not finished.  This is why
we add the tree for ``white'' to the derivation by the
\emph{adjunction} operation.  This makes the derivation syntactically
and semantically complete, and we are done.

\begin{figure}
  \centering
  xxxx
  \caption{Generating ``The white rabbit sleeps'' as a planning problem.}
  \label{fig:white-rabbit-as-planning}
\end{figure}

This process has clear parallels to planning: We manipulate a state by
applying actions in order to achieve a goal.  We can make this
connection precise by translating the SPUD problem into a planning
problem; the relevant excerpt of this problem is shown in
Fig.~\ref{fig:white-rabbit-as-planning}.  Each action in this planning
problem corresponds to the addition of a single elementary tree to the
derivation; the first parameter of the action is a node name in the
derivation tree, and the further parameters stand for the individuals
to which the semantic roles will be instantiated.  The syntactic
preconditions and effects are encoded using $\mathsf{open}$ and
$\mathsf{canadjoin}$ literals; the status of each referring expression
is tracked using $\mathsf{distractor}$ literals.  Notice that the
action effects contain terms of the form $\mathsf{subj}(u)$, which
construct new node names.  In order to represent the planning problem
in PDDL, these terms can be eliminated by estimating an upper bound
$n$ for the plan length, making $n$ copies of each action, ensuring
that copy $i$ can only be applied in step $i$, and replacing the term
$\mathsf{subj}(u)$ in an action copy by the constant
$\mathsf{subj}_i$.

In the example, the following plan solves the planning problem:
\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
\end{enumerate}

The grammatical derivation in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
sentence, can be systematically reconstructed from this plan.
Therefore we can solve the sentence generation problem via the detour
through planning and bring current search heuristics for planning to
bear on generation.
